---
title: "The Effect of Severe Weather on the Public Health and Economies of Communities in the United States"
synopsis: "Presented here is a distillation of data from the U.S. National Oceanic and Atmoshperic 
Administration on the effect severe wheather has on population health and the economy. To determine
the overall effect of population health the number of fatalities and injuries is considered. The amount
of property damage and crop damage will be used to determin the effect of the weather on the economy. Since
the data was collected by many different people there are not consistent identifiers for the types of weather events, ie
'TORNADOS' vs 'TORNADO'. A clustering method based on the distance between names was used to group events together. It was determined that
tornados by far cause the greates harm to public health when considering either injury or death. Droughts casue more damage
to crops than any other weather event while flooding tends to cause the most property damage as well as the second most
damage to crops."
output:
  html_document:
    toc: true
---
## Synopsis
Presented here is a distillation of data from the U.S. National Oceanic and
Atmoshperic Administration on the effect severe wheather has on population
health and the economy. To determine the overall effect of population health the
number of fatalities and injuries is considered. The amount of property damage
and crop damage will be used to determine the effect of the weather on the
economy. Since the data was collected by many different people there are not
consistent identifiers for the types of weather events, ie 'TORNADOS' vs
'TORNADO'. A clustering method based on the distance between event names was
used to group events together. It was determined that tornados by far cause the
greatest harm to public health when considering either injury or death. Droughts
casue more damage to crops than any other weather event while flooding tends to
cause the most property damage as well as the second most damage to crops.

### Data Provenence
The data are from the U.S. National Oceanic and Atmospheric Administration's database. The data can be downloaded [here](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2) and a description of the data can be found [here](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf)

### R Libraries and Global Settings
```{r, eval=TRUE, echo=TRUE, cache=TRUE}
library(knitr)
library(readr)
library(dplyr)
library(tidyr)
library(lubridate)
library(ggplot2)
library(scales)
library(stringdist)

opts_chunk$set(message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE)
```

# Data Processing

Download and read in the data.
```{r, cache=TRUE}
dataLink <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
tmpData <- tempfile()
download.file(dataLink, tmpData)
unprocessed <- read_csv(bzfile(tmpData), progress=FALSE)
unlink(tmpData)
```

## Merging the Event Types

There are `r length(unique(unprocessed$EVTYPE))` unique event types in the data
set. Many of them can be merged. For example 'AVALANCE' is probably a
misspelling of the event type 'AVALANCHE' or the event types 'RIP CURRENT' and
'RIP CURRENTS' are likely the same event type. Below is the code used to merge
event types. Event types are merged
based on clustering by similarity in their name (Jaro-Winker
distance[https://en.wikipedia.org/wiki/Jaro%E2%80%93Winkler_distance]) and the
event that represents the cluster is the one that was reported the most times in
the original data. This is a best effort approach to merging the events and is
not perfect, for example 'TSUNAMI' and 'BEACH FLOOD', are not likely to be
clustered together even though they are simular events. Only events that had distances in
the lower 10% were clustered together, so that only the most similar event names
were clustered together. However, the method of clustering (or not clustering at
all) did not have a large impact on the final statistics of the top results.

```{r, clusterEvents, cache=TRUE}

chooseEventRepresentitive <- function(events, eventCounts)
{
	filter(eventCounts, EVTYPE %in% events) %>% 
			arrange(desc(count)) %>% 
			select(EVTYPE) %>% 
			slice(1) %>% 
			as.character()
}
eventCounts <- group_by(unprocessed, EVTYPE) %>% dplyr::summarize(count=length(EVTYPE))
eventTypes <- select(unprocessed, EVTYPE) %>% unique() %>% arrange(EVTYPE) %>% tail(-1)
event.dist <- stringdistmatrix(gsub("[^A-Z]", "", toupper(eventTypes$EVTYPE)), method="jw")
tenth <- quantile(event.dist, 0.1)
eventTypes$cluster <- cutree(hclust(event.dist), h=tenth)

clusteredEvents <- group_by(eventTypes, cluster) %>% 
		dplyr::summarize(eventGroup=chooseEventRepresentitive(EVTYPE, eventCounts)) %>%
    	merge(eventTypes) %>%
    	select(-cluster) %>%
    	merge(unprocessed, by="EVTYPE") %>%
    	select(eventGroup, EVTYPE, FATALITIES, INJURIES, PROPDMG, PROPDMGEXP, CROPDMG, CROPDMGEXP)

```

## Totaling the Casualties
After the events have been clustered, totaling the fatalities
and injuries related to weather events is straight forward. The code below simply sums
the fatalities and injuries separately for each event.
```{r, sumCasualties, cache=TRUE}
casualties <- group_by(clusteredEvents, eventGroup) %>%
		dplyr::summarize(Fatalities=sum(FATALITIES), Injuries=sum(INJURIES)) %>%
		gather(type, amount, Fatalities, Injuries)
```

The column for representing the scale of damage done to property ('PROPDMGEXP')
and crops ('CROPDMGEXP') is not very consistant or straightforward. Some of the data
have been removed since it is not clear what the meaning of the 'PROPDMG' or
'CROPDMG' is. However, they have been dropped independently, that is if the
'CROPDMG' column is understandable but the 'PROPDMG' is not then that data point
is used in the final analysis for crop damage but not for property damage. To
sum the damages we first filter out data that we are unclear about and then
format the damage columns so that we can sum the total property and crop
(separately) for each event.

## Totaling the Damages
```{r, sumDamage, cache=T}
dropDmgExp <- c("?", "+", "-")

mutateDmgExp <- function(dmgExp)
{
	dmgExp <- ifelse(tolower(dmgExp) == "h", "2", dmgExp)
	dmgExp <- ifelse(tolower(dmgExp) == "k", "3", dmgExp)
	dmgExp <- ifelse(tolower(dmgExp) == "m", "6", dmgExp)
	dmgExp <- ifelse(tolower(dmgExp) == "b", "9", dmgExp)
	as.numeric(dmgExp)
}

colsep <- paste(sample(letters, 10), collapse="") #unlikely to collide with anything meaningful

damages <- select(clusteredEvents, eventGroup, PROPDMG:CROPDMGEXP) %>%
		dplyr::mutate(Property=paste(PROPDMG, PROPDMGEXP, sep=colsep), Crop=paste(CROPDMG, CROPDMGEXP, sep=colsep)) %>%
		select(eventGroup, Property, Crop) %>%
		gather(dmg.type, value, Property, Crop) %>%
		separate(value, into=c("dmg", "dmg.scale"), sep=colsep) %>%
		dplyr::mutate(dmg.scale=mutateDmgExp(dmg.scale), abs.dmg=as.numeric(dmg)*(10^dmg.scale)) %>%
		group_by(eventGroup, dmg.type) %>%
		dplyr::summarize(total.damage=round(sum(abs.dmg, na.rm=TRUE)/(10^9), 2))
```

# Results

## Weather Events Effect on Population Health

```{r, cache=TRUE}
top <- 15
topFatal <-  filter(casualties, type == "Fatalities") %>% 
    arrange(desc(amount)) %>% 
    head(top)
topInjury <- filter(casualties, type == "Injuries") %>% 
    arrange(desc(amount)) %>% 
    head(top)
fatalityRange <- filter(casualties,! eventGroup %in% topFatal$eventGroup) %>% select(amount) %>% range()
injuryRange <- filter(casualties,! eventGroup %in% topInjury$eventGroup) %>% select(amount) %>% range()
incommon <- intersect(topFatal$eventGroup, topInjury$eventGroup) %>% length()
# These next four lines allow for the plot to be ordered by highest amount independently
# across the facets
topBoth <- rbind(topInjury, topFatal) %>% arrange(amount, type) %>% as.data.frame()
topBoth$t <- as.numeric(factor(topBoth$type))
topBoth <- plyr::ddply(topBoth, .(type, eventGroup), transform, x=paste(c(rep(' ', t-1), eventGroup), collapse=''))
topBoth$x = factor(topBoth$x, levels=topBoth[order(topBoth$amount), 'x'])
ggplot(data=topBoth, aes(x=x, y=amount)) +
    facet_wrap(~type, scales="free", ncol=1) + 
    geom_bar(stat="identity") + 
    coord_flip() +
    xlab("Event Type") + 
    ylab("Total Amount") + 
    theme_bw()
```
The plot above shows the top `r top` events with respect fatalities and injuries (separately).
We can see from this plot that torandos have caused far more 
fatalities and injuries than the other events. For the other 
`r prettyNum(nrow(casualties) - top, big.mark=",", scientific=FALSE)` 
events the total fatalities range from `r fatalityRange[1]` to 
`r prettyNum(fatalityRange[2], big.mark=",", scientific=FALSE)` and the total 
injuries range from `r injuryRange[1]` to
`r prettyNum(injuryRange[2], big.mark=",", scientific=FALSE)`. There are
`r incommon` events in common amoung the top `r top` for fatalities and
injuries.


## Fiscal Effect of Weather Events
```{r, cache=TRUE}
top <- 15
topProp <-  filter(damages, dmg.type == "Property") %>% 
    arrange(desc(total.damage)) %>% 
    head(top)
topCrop <- filter(damages, dmg.type == "Crop") %>% 
    arrange(desc(total.damage)) %>% 
    head(top)
propDmgRange <- filter(damages,! eventGroup %in% topProp$eventGroup) %>% 
		as.data.frame() %>% 
		select(total.damage) %>% 
		range()
cropDmgRange <- filter(damages,! eventGroup %in% topCrop$eventGroup) %>% 
		as.data.frame() %>% 
		select(total.damage) %>% 
		range()
incommon.dmg <- intersect(topProp$eventGroup, topCrop$eventGroup) %>% length()
# These next four lines allow for the plot to be ordered by highest amount independently
# across the facets
topBoth <- rbind(topProp, topCrop) %>% arrange(total.damage, dmg.type) %>% as.data.frame()
topBoth$t <- as.numeric(factor(topBoth$dmg.type))
topBoth <- plyr::ddply(topBoth, .(dmg.type, eventGroup), transform, x=paste(c(rep(' ', t-1), eventGroup), collapse=''))
topBoth$x = factor(topBoth$x, levels=topBoth[order(topBoth$total.damage), 'x'])
ggplot(data=topBoth, aes(x=x, y=total.damage)) +
    facet_wrap(~dmg.type, scales="free", ncol=1) + 
    geom_bar(stat="identity") + 
    coord_flip() +
    xlab("Event Type") + 
    ylab("Total Amount (in billions of dollars") + 
    theme_bw()
```

From the plot above we can see that droughts have the greatest fiscal
inpact on crops and are not even in the top `r top` event types for property
damage, while flooding by far causes the most property damage and the second
most crop damage. It is worth highlightling that property damages are in
general much higher (about 10x) than crop damages. For the r
`r prettyNum(nrow(damages) - top, big.mark=",", scientific=FALSE)` 
events the total crop damages range from `r cropDmgRange[1]` to 
`r prettyNum(cropDmgRange[2], big.mark=",", scientific=FALSE)` billion dollars
and the total injuries range from `r propDmgRange[1]` to
`r prettyNum(propDmgRange[2], big.mark=",", scientific=FALSE)` billion dollars.
There are `r incommon.dmg` events in common amoung the top `r top` most damaging
events.

