---
author : "John Willey"
date : "August 11th 2017"
title : "Simple Random Forest Modeling Application"
output : ioslides_presentation
---

## Application Overview

The application has one goal...to build a random forest model of a dataset.

Features:

* Users can select from two available data sets!
* You decide what outcome to model.
* Pick your features.
* Want to hold some data to test your model? No problem.
* Download your results.

## Building Your Model with the Application
Selecting the options below, as in the screen shot, is the same as running the code on the next slide.

```{r, echo=FALSE}
library(png)
library(grid)
screenshot <- readPNG("/Users/jwilley44/Desktop/data_app_screenshot.png")
grid.raster(screenshot)
```

## R Code Equivalent

```{r, echo=TRUE, warning=FALSE, message=FALSE, }
library(caret)
library(randomForest)
library(dplyr)
set.seed(1234)
iris.train <- iris
tr <- createDataPartition(iris.train$Species, p=0.8)
iris.train$set <- 1:nrow(iris) %in% tr$Resample1
iris.train <- mutate(iris.train, set=ifelse(set, "training", "test"))
rf_model <- randomForest(
    Species ~ Sepal.Length + Sepal.Width + Petal.Length + Petal.Width, 
    data=subset(iris.train, set=="training"))
```

## Results
We can see that the running the R code gives the same result as the application.
```{r}
rf_model
```

