---
title: "Tooth Growth by Dose and Supplement"
author: "John Willey"
output: pdf_document
---

```{r, eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE}
library(knitr)
library(ggplot2)
library(dplyr)
library(datasets)
opts_chunk$set(message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE)

pvalues <- data.frame(data_names=c(), p.value=c())
addPvalue <- function(pvalues, dataname, pvalue) rbind(pvalues, data.frame(data_names=dataname, p.value=pvalue))
```

# Overview
The analysis below aims to use the central limit theorem to test that mean of lenghts for given supplements and doses are different.

# Summary of Data
A quick summary of the tooth growth data is below, showing the mean and standard deviation for the lenghts for each supplement and dose, as well as the number of observations. The histogram below shows the distributions of the lengths with doses horizontally and supplements vertically.
```{r, echo=FALSE}
kable(group_by(ToothGrowth, supp, dose) %>% 
          summarize(mean.length=mean(len), sd.length=round(sd(len), 2), min.length=min(len), max.length=max(len), observations=length(len)))
```
```{r, echo=FALSE, fig.align='center', fig.height=3, fig.width=6}
qplot(data=ToothGrowth, x=len) + facet_grid(supp~dose) + theme_bw()
```
# Comparison of Lengths Accross Doses and Supplements
From the histograms above we can see that the lenghts under either supplement increase when the dose increases, while the distributions do not look qualitatively different between supplements (for a given dose). In the anlyses below the function "brief_t.test" just uses R's  "t.test" with the "alternative" option set to greater and it only prints out the p-value and the confidence interval.

```{r, echo=FALSE}
brief_t.test <- function(x, y, d1, d2)
{
    results <- t.test(x, y, alternative="greater")
    pvalues <<- addPvalue(pvalues, paste(d1, "and", d2), results$p.value)
    print(paste("p-value =", results$p.value))
    print(paste("95% confidence interval: ", results$conf.int[1], "to", results$conf.int[2]))
}
```

## Comparison Between Doses
In the analyses below variables representing different statistics and quanties will be denoted with a subscript of the dose to which they pertain.
```{r, echo=F}
lengths_0.5 <- filter(ToothGrowth, dose == 0.5)$len
lengths_1.0 <- filter(ToothGrowth, dose == 1.0)$len
lengths_2.0 <- filter(ToothGrowth, dose == 2.0)$len
```
### Dose of 0.5 vs 1.0
Statement of hypothesis:  
$H_0$: $\mu_{0.5} = \mu_{1.0}$  
$H_{1.0>0.5}$: $\mu_{1.0} > \mu_{0.5}$  
The mean value of the lengths at dose 1.0 is larger than the mean of the lengths at a dose of 0.5.  
```{r, message=T}
brief_t.test(lengths_1.0, lengths_0.5, "1.0", "0.5")
```
### Dose of 0.5 vs 2.0
Statement of hypothesis:  
$H_0$: $\mu_{0.5} = \mu_{2.0}$  
$H_{2.0>0.5}$: $\mu_{2.0} > \mu_{0.5}$  
The mean value of the lengths at dose 2.0 is larger than the mean of the lengths at a dose of 0.5.

```{r, message=T}
brief_t.test(lengths_2.0, lengths_0.5, "2.0", "0.5")
```
### Dose of 1.0 vs 2.0
Statement of hypothesis:  
$H_0$: $\mu_{1.0} = \mu_{2.0}$  
$H_{2.0>1.0}$: $\mu_{2.0} > \mu_{1.0}$  
The mean value of the lengths at dose 2.0 is larger than the mean of the lengths at a dose of 1.0.

```{r}
brief_t.test(lengths_2.0, lengths_1.0, "2.0", "1.0")
```
## Comparison Between Supplements
```{r, echo=FALSE}
lengths_OJ <- filter(ToothGrowth, supp == "OJ")$len
lengths_VC <- filter(ToothGrowth, supp == "VC")$len
```

Since we cannot tell which one will be greater, we perform the t-test with the alternative set to "two.sided" and therefore our hypothesis statement is:  
$H_0$: $\mu_{OJ} = \mu_{VC}$  
$H_{OJ > VC}$ : $\mu_{OJ} > \mu_{VC}$

```{r, message=T}
brief_t.test(lengths_OJ, lengths_VC, "OJ", "VC")
```

## Comparison of Doses and Supplements
The code to produce this plot  can be found in the appendix in section ["Compare All..."](#comp-all). Each tile in the plot is an experiment comparing two slices of the data. The hypothesis for each tile is of the form:  
$H_0$: $\mu_{x} = \mu_{y}$  
$H_{x>y}$: $\mu_{x} > \mu_{y}$  
Where $x$ and $y$ represent the x-axis and y-axis respectively.

```{r, echo=FALSE, eval=TRUE}
groupedLengths <- group_by(ToothGrowth, dose, supp) %>%
    do(lengths=.$len) %>%
    mutate(merge.col="merge.here")

df_t.test <- function(df)
{
    l1 <- df$lengths.x
    l2 <- df$lengths.y
    results <- t.test(l1[[1]], l2[[1]], alternative="greater")
    # Adds the global list of p-values
    pvalues <<- addPvalue(pvalues, paste(df$label.x[1], "and", df$label.y), results$p.value)
    data.frame(
        p.value=results$p.value, 
        lower=results$conf.int[1], 
        upper=results$conf.int[2],
	    result=paste(signif(results$p.value, 2), 
	             paste0("[", round(results$conf.int[1], 2), ", ", 
	                    round(results$conf.int[2], 2), "]" ), sep="\n"))
}

t.tests <- merge(groupedLengths, groupedLengths, by="merge.col") %>%
    mutate(label.x=paste(dose.x, supp.x), label.y=paste(dose.y, supp.y)) %>%
    group_by(label.x, label.y) %>%
    do(df_t.test(.)) %>%
    mutate(conclusion=ifelse(lower > 0 & p.value < 0.05, 
                             "Reject Null Hypothesis",
                             "Accept Null Hypothosis"))

ggplot(data=t.tests, aes(x=label.x, y=label.y)) + 
    geom_tile(aes(fill=conclusion), colour="black") + 
    geom_text(aes(label=result)) + 
    xlab("") + ylab("") + theme(legend.position="bottom", legend.title=element_blank())

```

# Conclusions
The central limit theorem allows us to ignore the underlying distribution of lengths when we consider the mean of lengths for the given slices of data. From the t-tests above we can resonably say that with 95% confidence that the mean lenght associated with larger doses are larger than those with smaller doses and that the mean length associated with suppliement "OJ" are larger than those associated with supplement "VC". However, from the p-values and the confidence intervals that a bigger difference in the means is associated with dose and not suppliement type. In fact the lower end of the confidence interval for "OJ" mean length larger than "VC" is fairly close to zero suggesting that there may not be a large difference between the two actual means. The plot showing all the comparisons echos what we learned from the previous tests, that higher doses have higher mean lengths than lower doses. However, we can see that for dose = 2.0, we cannot say that the means of the lengths for "OJ" and "VC" are different. Forty different hypothesese were tested in this analysis. Using the method of Bonferroni, the adjusted p-values (in [appendix](#adjust-p-values)) show that there were 5 cases in which the adjusted p-value is no longer considered sufficient.

#Apendix
## Code to Produce the Plot Comparing All Slices of Data{#comp-all}

```{r, echo=TRUE, eval=FALSE}
groupedLengths <- group_by(ToothGrowth, dose, supp) %>%
    do(lengths=.$len) %>%
    mutate(merge.col="merge.here")

df_t.test <- function(df)
{
    l1 <- df$lengths.x
    l2 <- df$lengths.y
    results <- t.test(l1[[1]], l2[[1]], alternative="greater")
    # Adds the global list of p-values
    pvalues <<- addPvalue(pvalues, paste(df$label.x[1], "and", df$label.y), results$p.value)
    data.frame(
        p.value=results$p.value, 
        lower=results$conf.int[1], 
        upper=results$conf.int[2],
	    result=paste(signif(results$p.value, 2), 
	             paste0("[", round(results$conf.int[1], 2), ", ", 
	                    round(results$conf.int[2], 2), "]" ), sep="\n"))
}

t.tests <- merge(groupedLengths, groupedLengths, by="merge.col") %>%
    mutate(label.x=paste(dose.x, supp.x), label.y=paste(dose.y, supp.y)) %>%
    group_by(label.x, label.y) %>%
    do(df_t.test(.)) %>%
    mutate(conclusion=ifelse(lower > 0 & p.value < 0.05, 
                             "Reject Null Hypothesis",
                             "Accept Null Hypothosis"))

ggplot(data=t.tests, aes(x=label.x, y=label.y)) + 
    geom_tile(aes(fill=conclusion), colour="black") + 
    geom_text(aes(label=result)) + 
    xlab("") + ylab("") + theme(legend.position="bottom", legend.title=element_blank())
```

## Adjusted P-values {#adjusted-p-values}

```{r, echo=FALSE, results='asis'}
pvalues$adjusted.p.value <- signif(p.adjust(pvalues$p.value, method="bonf"), 3)
pvalues$p.value <- signif(pvalues$p.value, 3)
kable(filter(pvalues, p.value < 0.05, adjusted.p.value < 0.05),
      caption="Significant adjusted p-value")
kable(filter(pvalues, p.value < 0.05, adjusted.p.value >= 0.05),
      caption="Significant p-value, Not significant adjusted p-value")
kable(filter(pvalues, p.value >= 0.05),
      caption="Not significant p-value")    
```