# Reading from MySQL

## Access the UCSC Genomics database in R
```{r}
library(RMySQL)
ucscDb <- dbConnect(MySQL(), user="genome", host="genome-mysql.cse.ucsc.edu")
```

## Resources
sqldf package for sql operations on dataframes  
https://cran.r-project.org/web/packages/RMySQL/RMySQL.pdf  
https://www.r-bloggers.com/mysql-and-r/  
http://www.pantz.org/sofware/mysql/sqlcommands.html  

# Reading data from HDF5

## Example
```{r, eval=TRUE}
library(rhdf5)
created <- h5createFile("example.h5")
h5createGroup("example.h5", "foo")
h5createGroup("example.h5", "baa")
h5createGroup("example.h5", "foo/foobaa")
h5ls("example.h5")

A <- matrix(1:10, nr=5, nc=2)
h5write(A, "example.h5", "foo/A")
h5ls("example.h5")
readA <- h5read("example.h5", "foo/A")
readA

h5write(c(12, 13, 14), "example.h5", "foo/A", index=list(1:3, 1))
h5read("example.h5", "foo/A")

```

## Resources
https://www.hdfgroup.org/  
https://www.bioconductor.org/packages/release/bioc/vignettes/rhdf5/inst/doc/rhdf5.pdf  

# Reading data by Web Scraping

## Example
```{r, eval=TRUE}
con  <- url("https://scholar.google.com/citations?user=HI-I6C0AAAAJ")
htmlCode <- readLines(con, warn=FALSE)
close(con)
library(XML)
html <- htmlTreeParse(htmlCode, useInternalNodes=T)
xpathSApply(html, "//title", xmlValue)
xpathSApply(html, "//td[@id='col-citeby']", xmlValue)

library(httr)
html2 <- GET("https://scholar.google.com/citations?user=HI-I6C0AAAAJ")
content2 <- content(html2, as="text")
parsedHtml <- htmlParse(content2, asText=TRUE)
xpathSApply(parsedHtml, "//title", xmlValue)

```

## Example Authentication
```{r, eval=FALSE}
webPage <- GET("www.theurl.com/blorg", authenticate("user", "password"))
```

## Resources

http://en.wikipedia.org/wiki/Web_scraping  
https://cran.r-project.org/web/packages/httr/httr.pdf  
https://www.r-bloggers.com/search/web%20scraping/  

# Reading data from API

## Example reading from Twitter
```{r eval=FALSE}
myapp <- oauth_app("twitter", key="yourConsumerKey", secret="yourConsumerSecret")
sig <- sign_oauth1.0(myapp, token="yourToken", token_secret="yourTokenSecret")
homeTl <- GET("https:/api.twitter.com/1.1/statuses/home_timeline.json", sig)
content(homeTL)
library(jsonlite)
jsonContent <- fromJSON(toJSON(content(homeTL)))
```

## Resources
https://github.com/hadley/httr/tree/master/demo  

# Reading from other sources

## foreign package
read.arf (Weka)  
read.dta (Stat)  
read.mtp (Minitab)  
read.octave (Octave)  
read.spss (SPSS)  
read.xport (SAS)  

https://cran.r-project.org/web/packages/foreign/foreign.pdf  

## Other database pacakges
###Postgress
https://cran.r-project.org/web/packages/RPostgreSQL/index.htm  l
https://www.r-bloggers.com/getting-started-with-postgresql-in-r/  
https://www.r-bloggers.com/using-postgresql-in-r-a-quick-how-to/  
### RODBC
https://cran.r-project.org/web/packages/RODBC/RODBC.pdf  
https://cran.r-project.org/web/packages/RODBC/vignettes/RODBC.pdf  
### MongoDb  
https://cran.r-project.org/web/packages/mongolite/vignettes/intro.html  

## Images
jpeg https://cran.r-project.org/web/packages/jpeg/jpeg.pdf  
readbitmap https://cran.r-project.org/web/packages/readbitmap/readbitmap.pdf  
png https://cran.r-project.org/web/packages/png/png.pdf  
EBImage https://bioconductor.org/packages/release/bioc/html/EBImage.html  

## GIS data
rdgal https://cran.r-project.org/web/packages/rgdal/rgdal.pdf  
rgeos https://cran.r-project.org/web/packages/rgeos/rgeos.pdf  
raster https://cran.r-project.org/web/packages/raster/raster.pdf  

## Music data
tuneR https://cran.r-project.org/web/packages/tuneR/tuneR.pdf  
seewave https://cran.r-project.org/web/packages/seewave/seewave.pdf  
