# Cleaning the data
In this directory you can run
```
./run_analysis.R
```
This will create a directory called tidyData in this directory with two
files:    
tidyData.tsv contains the mean and standard deviation of all the data points  
tidyData.summarized.tsv contains the average values for the mean and standard deviation for the data  
SeeCodeBook.md for more information on the format of those files

## Step 1: Setting up the environment
Set the appropriate location of
the data.

```{r eval=TRUE}
originalDataDir <- "originalData"
source("Rcode/tidyData.R")
setInitialVariables(originalDataDir)
```

## Step 2: Load the data
The commands below load the data into a data frame, label the test and training
sets, combine the data.
```{r eval=TRUE}
train.data <- getTrainFeatureData(originalDataDir)
test.data <- getTestFeatureData(originalDataDir)
all.data <- rbind(train.data, test.data) %>% tbl_df()
print(all.data)
```

## Step 3: Tidy the data
Tidys the data by separting out the feature column from above.
```{r eval=TRUE}
tidy.data <- tidyFeatureData(all.data)
print(tidy.data)
```
See CodeBookd.md for a description of the data

## Step 4: Summarize the data
Summarize the data over set, activity, metric, measurement, direction by taking
the average values. Spread out the measurement column into two columns, mean and
standard deviation.
```{r eval=TRUE}
summarized.data <-
summarizeFeatureData(tidy.data)
print(summarized.data)
```
See CodeBookd.md for a description of the data